LEGWORK: LISA Evolution and Gravitational Wave Orbit Kit

### 2021-11-10 src/ms.tex
hack the showyourwork colours to match scheme
update repo to match Overleaf

% define colour from xcolor
\definecolor{SeaGreen}{RGB}{63, 188, 157}

@@

\begin{abstract}
    We present \lw{}, a Python package for making predictions about stellar-origin gravitational wave sources and their detectability in LISA and other space-based gravitational wave detectors. \lw{} can be used to evolve sources over time, calculate gravitational wave strains, compute signal-to-noise ratios and visualise the results. In this paper we give an overview of the structure and functionality of the package, as well as highlight several optimisations that make it particularly efficient. We demonstrate the various capabilities of \lw{} through a series of example use cases. These begin with a basic example of calculating the signal-to-noise ratio of a source but also include using \lw{} to calculate horizon distances and make comparisons between gravitational-wave detectors. We additionally explain in detail the derivation of the equations behind \lw{} in order to clarify some conflicts currently present in the literature.
\end{abstract}

@@

The planned space-based gravitational wave detector LISA (Laser Interferometer Space Antenna) will present an entirely new view of gravitational waves by focusing on lower frequencies ($10^{-5} < f / \unit{Hz} < 10^{-1}$) than ground-based detectors. This will enable the study of the gravitational wave mergers of supermassive black holes that are undetectable with ground-based detectors \citep[e.g.][]{Begelman+1980, Klein+2016}. However, this frequency regime is also of interest for the detection of local stellar-mass binaries during their inspiral phase. LISA is expected to detect Galactic stellar-origin compact objects, ranging from the numerous double white dwarf (WDWD) population, to the rare but loud double black hole (BHBH) population.

@@

We believe that the large renewed interest in LISA and the stellar-origin sources it may detect will lead to many more studies in the near future that would need similar computations. This leads to a significant amount of redundancy which, at best results in extra work for each individual and at worst leads to an increased chance of introducing mistakes and inconsistencies when coding up the necessary expressions.

\lw{} is a Python package designed to streamline the process of making predictions of LISA detection rates for stellar-origin binaries such that it is fast, reliable and simple. With \lw{} one can evolve the orbits of a binary or a collection of binaries and calculate their strain amplitudes for any range of harmonics. One can compute the sensitivity curve for LISA or other future gravitational wave detectors (e.g. TianQin's curve, or that of a custom instrument) and use it to compute the signal-to-noise ratio of a collection of sources. Furthermore, \lw{} provides tools to visualise all of the results with easy-to-use plotting functions.

Specifically, we implement the expressions by \citet{Peters+1963} and \citet{Peters+1964} for the evolution of binary orbits, equations for the strain amplitudes and signal-to-noise ratios of binaries from various papers \citep[e.g.][]{Flanagan+1998, Finn+2000, Barack+2004, Moore+2015} and approximations for the LISA and TianQin sensitivity curves given in \citet{Robson+2019} and \citet{Huang+2020} respectively.

@@

The paper is organised as follows. In Section~\ref{sec:LEGWORK-overview} we give an overview of the capabilities of the \lw{} package and its various modules. In Section~\ref{sec:example-uses} we outline some example use cases of \lw{} to demonstrate its use. We detail the derivations of the equations relevant for \lw{} in Section~\ref{sec:derivations}. Finally, in Section~\ref{sec:summary} we conclude and summarise our work.

@@

The \lw{} package is composed of seven modules that each focus on a particular aspect of calculations useful for LISA gravitational wave sources. In Fig.~\ref{fig:package_overview}, we illustrate the general structure of the package with its various modules. The source module is the central module of the package and provides an simple interface to the functions in the rest of the modules. For more complex analyses, users may want to interact directly with individual modules, particularly those in the top row of Fig.~\ref{fig:package_overview} as they comprise the core functionality of \lw{}. Below we explain the capabilities of each module in detail.

\lwModLink{evol} deals with the orbital evolution of a binary due to gravitational waves. It includes functions for computing the merger times of both circular and eccentric binaries. In addition, you can use this module to evolve binary orbit parameters forward in time with any number or arrangement of timesteps. We discuss the relevant equations in Sec.~\ref{sec:deriv_evol}.

\lwModLink{strain} contains two functions that compute a binary's gravitational wave strain and characteristic strain amplitude respectively. Each of these functions is capable of computing the strain for a 3-dimensional array of binaries at any number of timesteps and evaluated at any number of harmonics. We discuss the relevant equations in Sec.~\ref{sec:deriv_strain}.

@@

\lwModLink{source} provides a direct and simple interface to the functions in other modules through the \texttt{Source} Class. You can instantiate this Class with an array of sources and use it compute their strains or signal-to-noise ratios directly. Moreover, depending on the user's choice of allowed gravitational wave luminosity error, the Class dynamically decides on the number of harmonics needed to capture the full signal of each binary and at what eccentricity to no longer consider a binary circular. This Class also provides a quick means of evolving the sources, visualising the parameters of each source and allows you to plot the binaries on the sensitivity curve.

@@

@@

Moreover, for increasingly eccentric sources, gravitational waves are emitted in an increasing number of higher harmonics. Although the total SNR is formally calculated as a the sum of the SNR over an infinite number of harmonics, in practice it is sufficient to only consider a subset. In the \lwModLink{source} module, the user can provide \texttt{gw\_lum\_tol}, a maximum allowable tolerance for the accuracy of the gravitational wave luminosity. Given this tolerance, \lw{} automatically calculates the required number of harmonics to satisfy this tolerance and thus minimise the computation time.

@@

@@

Let us compute the horizon distance for a grid of chirp masses and orbital frequencies. First, we can recall that the signal-to-noise ratio of a source is inversely proportional to its distance from LISA and so we can find the horizon distance, $D_{\rm hor}$ as
\begin{equation}{}
    \label{eq:snr_to_hor_dist}
    D_{\rm hor} = \frac{\rho(D)}{\rho_{\rm detect}} \cdot D,
\end{equation}
where $\rho(D)$ is the signal-to-noise ratio as some distance $D$ and $\rho_{\rm detect}$ is the threshold above which we consider a source to be detectable. For the purpose of this example we will set $\rho_{\rm detect} = 7$. Therefore, we can start by finding the signal-to-noise ratio for each source at an arbitrary distance and convert that to a horizon distance.

This can be most efficiently accomplished using \lw{}'s source class. Although the source class only takes one-dimensional lists of sources, one can flatten a grid of chirp masses and orbital frequencies to make the input work. \lw{} can then compute the merger times and signal-to-noise ratio of each source with only two lines of code. These results can then be reshaped to match the original grid of chirp masses and frequencies.

We convert these SNRs to horizon distances using Eq.~\ref{eq:snr_to_hor_dist} and plot the result in Figure~\ref{fig:horizon_distance}. The shape of the LISA sensitivity curve is clearly reflected in this plot of the horizon distance. This is because circular sources emit the vast majority of their single at a single frequency and thus every feature in the sensitivity curve has a strong effect on the SNR (and therefore horizon distance) for circular sources. We also see that once the merger time of a source is shorter than the LISA mission length (shown by the black dotted line), the horizon distance sharply decreases. This is because if a source merges before the LISA mission concludes, it has less time to accumulate signal and thus has a lower SNR and horizon distance.

@@

We see two effects in the signal-to-noise ratio here. First, increasing the eccentricity from essentially circular to $e = 0.6$ results in a higher signal-to-noise ratio ($\rho=31.7 \to \rho=50.2$). This is because an eccentric binary has enhanced energy emission via gravitational waves \citep{Peters+1963}. This means that an eccentric binary will not only inspiral faster than an otherwise identical circular binary, but also will always have a stronger gravitational wave strain. The enhancement factor and its exact dependence on eccentricity is discussed in more detail Section~\ref{sec:derivations} (specifically Eq.~\ref{eq:eccentricity_enhancement_factor}).

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=\columnwidth]{figures/detector_sc_compare.pdf}
    \caption{The strain spectral density of the LISA detector (with different specifications) and the TianQin detector. We show the LISA curve for three different mission lengths and once with an approximate response function.}
    \label{fig:detector_sc_compare}
\end{figure}

@@

Note that \lw{} could also be used in this way to find the SNR of any system in which the orbital evolution is know. Thus for a triple system or a binary experiencing gas drag (as long the evolution of the eccentricity and orbital frequency is known) \lw{} is entirely capable of calculating the SNR evolution.

@@

% \subsection{SNR calculation for verification binaries}

%\section{Benchmarks}\label{sec:speed}
%A section to show off \lw{}'s speed.

A \docsIcon{} symbol in an equation will link to the relevant \lw{} documentation for the implementation of that equation. These derivations are also given in more detail in the \lw{} documentation \href{https://legwork.readthedocs.io/en/latest/notebooks/Derivations.html}{here}, where we show each of the intervening steps in more detail.

>>>>

% define colour from xcolor
\definecolor{SeaGreen}{RGB}{63, 188, 157}
\colorlet{sywScriptUpToDate}{SeaGreen}

@@

\begin{abstract}
    We present \lw{}, a Python package for making predictions about stellar-origin gravitational wave sources and their detectability in LISA and other space-based gravitational wave detectors. \lw{} can be used to evolve sources over time, calculate gravitational wave strains, compute signal-to-noise ratios and visualise the results. In this paper we give an overview of the structure and functionality of the package, as well as highlight several optimisations that make it particularly efficient for use in population studies which can contain tens-of-millions of sources. We demonstrate the various capabilities of \lw{} through a series of example use cases. These begin with a basic example of calculating the signal-to-noise ratio of a source but also include using \lw{} to calculate horizon distances and make comparisons between gravitational-wave detectors. We additionally explain in detail the derivation of the equations behind \lw{} in order to clarify some conflicts currently present in the literature.
\end{abstract}

@@

The planned space-based gravitational wave detector LISA (Laser Interferometer Space Antenna) will present an entirely new view of gravitational waves by focusing on lower frequencies ($10^{-5} < f / \unit{Hz} < 10^{-1}$) than ground-based detectors. This will enable the study of many source classes that are undetectable with ground-based detectors including mergers of supermassive black holes that are undetectable with ground-based detectors \citep[e.g.][]{Begelman+1980, Klein+2016, Bellovary2019}, extreme mass ratio inspirals \citep[e.g.,][]{Berti2006, Barack2007, Babak2017, Moore2017}, and cosmological GW backgrounds \citep[e.g.,][]{Bartolo2016, Caprini2016, Caldwell2019}. However, this frequency regime is also of interest for the detection of local stellar-mass binaries during their inspiral phase. LISA is expected to detect Galactic stellar-origin binaries containing combinations of white dwarfs, neutron stars, and black holes, ranging from the numerous double white dwarf population, to the rare but loud double black hole population.

@@

We believe that the large renewed interest in LISA and the stellar-origin sources it may detect will lead to many more studies in the near future that would need similar computations. This leads to a significant amount of redundancy which, at best results in extra work for each individual and at worst leads to an increased chance of introducing mistakes and inconsistencies when translating the necessary expressions to software.

\lw{} is a Python package designed to streamline the process of making predictions of LISA detection rates for stellar-origin binaries such that it is as fast, reliable and simple as possible. This goal makes \lw{} unique among other implementations of gravitational-wave tools in the literature, which focus on a more broad coverage of the gravitational-wave spectrum and source classes, rather than an optimised approach for certain sources. \citep[e.g.][]{Moore+2015,Yi2021}. With \lw{} one can evolve the orbits of a binary or a collection of binaries and calculate their strain amplitudes for any range of frequency harmonics. One can compute the sensitivity curve for LISA or other future gravitational wave detectors (e.g. TianQin's curve, or that of a custom instrument) and use it to compute the signal-to-noise ratio of a collection of sources. Furthermore, \lw{} provides tools to visualise all of the results with easy-to-use plotting functions. Finally, \lw{} is fully tested to check for consistency in the derivations described below.

Specifically, we implement the expressions by \citet{Peters+1963} and \citet{Peters+1964} for the evolution of binary orbits due to the emission of gravitational waves, equations for the strain amplitudes and signal-to-noise ratios of binaries from various papers \citep[e.g.][]{Flanagan+1998, Finn+2000, Cornish2003, Barack+2004, Moore+2015} and approximations for the LISA and TianQin sensitivity curves given in \citet{Robson+2019} and \citet{Huang+2020} respectively.

@@

The paper is organised as follows. In Section~\ref{sec:LEGWORK-overview} we give an overview of the capabilities of the \lw{} package and its various modules. We detail the derivations of the equations relevant for \lw{} in Section~\ref{sec:derivations}. In Section~\ref{sec:example-uses} we outline some example use cases of \lw{} to demonstrate its use. Finally, in Section~\ref{sec:summary} we conclude and summarise our work.

@@

The \lw{} package is composed of seven modules that each focus on a particular aspect of calculations useful for LISA gravitational wave sources. In Fig.~\ref{fig:package_overview}, we illustrate the general structure of the package with each of its modules. The source module is the central module of the package and provides an simple interface to the functions in the rest of the modules. For more complex analyses, users may want to interact directly with individual modules, particularly those in the top row of Fig.~\ref{fig:package_overview} as they comprise the core functionality of \lw{}. Below we explain the capabilities of each module in detail.

\lwModLink{evol} handles the orbital evolution of a binary due to the emission of gravitational waves. It includes functions for computing the merger times of both circular and eccentric binaries. In addition, you can use this module to evolve binary orbit parameters forward in time with any number or arrangement of timesteps. We discuss the relevant equations in Sec.~\ref{sec:deriv_evol}.

\lwModLink{strain} contains two functions that compute a binary's gravitational wave strain and characteristic strain amplitude respectively. Each of these functions is capable of computing the strain for a 3-dimensional array of binaries at any number of timesteps and evaluated at any number of frequency harmonics. We discuss the relevant equations in Sec.~\ref{sec:deriv_strain}.

@@

\lwModLink{source} provides a direct and simple interface to the functions in other modules through the \texttt{Source} Class. You can instantiate this Class with an array of sources and use it compute their strains or signal-to-noise ratios directly. Moreover, depending on the user's choice of allowed gravitational wave luminosity error, the Class dynamically decides on the number of frequency harmonics needed to capture the full signal of each binary and at what eccentricity to no longer consider a binary circular. This Class also provides a quick means of evolving the sources, visualising the parameters of each source and allows you to plot the binaries on the sensitivity curve.

@@

To ensure stability with physical units, all quantities included in \lw{} use the \texttt{astropy.units} module \citep{AstropyCollaboration+2013, AstropyCollaboration+2018}. This means that all inputs to \lw{} can be given in the units of the user's choice and will be automatically converted. Furthermore, all of the source code in \lw{} is fully tested with continuous integration in the \lw{} GitHub repository. We employ several unit tests to ensure consistency between each of the use cases described below. For example, we require that the SNR calculation for circular and stationary binaries produces consistent output whether \lw{} uses the stationary and circular approximations or not. Similarly, we verify that the antenna patterns described below produce the expected values when averaged over source positions, inclinations, and polarisations.

@@

Moreover, for increasingly eccentric sources, gravitational waves are emitted in an increasing number of higher frequency harmonics. Although the total SNR is formally calculated as a the sum of the SNR over an infinite number of harmonics, in practice it is sufficient to only consider a subset. In the \lwModLink{source} module, the user can provide \texttt{gw\_lum\_tol}, a maximum allowable tolerance for the accuracy of the gravitational wave luminosity. Given this tolerance, \lw{} automatically calculates the required number of harmonics to satisfy this tolerance and thus minimise the computation time.

@@

A \docsIcon{} symbol in an equation will link to the relevant \lw{} documentation for the implementation of that equation, which additionally contains a link to exact code used to reproduce the equation. These derivations are also given in more detail in the \lw{} documentation \href{https://legwork.readthedocs.io/en/latest/notebooks/Derivations.html}{here}, where we show each of the intervening steps in more detail.

For eccentric binaries, we need to consider all frequency harmonics of gravitational-wave emission. These are defined such that the $n^{\rm th}$ harmonic frequency is
\begin{equation}{}
    f_{n} \equiv n \cdot f_{\rm orb}.
    \label{eq:nth_freq}
\end{equation}
It will be important to know the relative gravitational-wave power radiated into the $n^{\rm th}$ harmonic for a binary with eccentricity $e$ for the following strain and SNR calculations. This is given by \citep[][Eq.\,20]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_g}}
    \begin{aligned}
        g(n, e) = \frac{n^{4}}{32} & \Big\{ \Big[ J_{n-2}(n e)-2 e J_{n-1}(n e)+\frac{2}{n} J_{n}(n e) \\
        &\left. +2 e J_{n+1}(n e)-J_{n+2}(n e)\Big]^{2}\right.\\
        &\left.\kern-\nulldelimiterspace\right.
            \!\!\begin{aligned}
            +\big(1-e^{2}\big)\Big[&J_{n-2}(n e)-2 J_{n}(n e) \\
            &+J_{n+2}(n e)\Big]^{2} \\
            \end{aligned}\\
        &+\frac{4}{3 n^{2}}\Big[J_{n}(n e)\Big]^{2}\Big\},
    \end{aligned}
    \label{eq:g(n,e)}
\end{equation}
where $J_{n}(v)$ is the Bessel function of the first kind. Thus, the \textit{sum} of $g(n, e)$ over all harmonics gives the factor by which the gravitational-wave emission is stronger for a binary of eccentricity $e$ over an equivalent circular binary. This enhancement factor is \citep[][Eq.\,17]{Peters+1963}
\begin{equation}{\lwFuncURL{utils}{peters_f}}
    F(e) \equiv \sum_{n = 1}^{\infty} g(n, e) = \frac{1 + \frac{73}{24} e^2 + \frac{37}{96} e^4}{(1 - e^2)^{7/2}}.
    \label{eq:eccentricity_enhancement_factor}
\end{equation}
A useful rule of thumb is that $F(0.5) \approx 5.0$, or in words, a binary with eccentricity $0.5$ loses energy to gravitational waves at 5 times the rate of an equivalent circular binary.

\subsection{Orbital evolution {\normalfont (\texorpdfstring{\lwModLink{evol}}{evol})}}\label{sec:deriv_evol}
\subsubsection{Circular binaries}
For a circular binary, the orbital evolution due to gravitational-wave emission can be calculated analytically, as the rate at which the separation of the binary shrinks is simply a function of its mass and the current separation \citep[][Eq.~5.6]{Peters+1964}
\begin{equation}{}
    \dv{a}{t}_{e = 0} = -\frac{\beta}{a^3},
\end{equation}
where the constant $\beta$ is defined as
\begin{equation}{\lwFuncURL{utils}{beta}}
    \beta(m_1, m_2) = \frac{64}{5} \frac{G^3}{c^5} m_1 m_2 (m_1 + m_2),
    \label{eq:beta_peters}
\end{equation}
where $m_1$ is the primary mass and $m_2$ is the secondary mass. This gives the semi-major axis of a circular binary as a function of time, $t$, as \citep[][Eq.~5.9]{Peters+1964}
\begin{equation}{\lwFuncURL{evol}{evol_circ}}
    a(t, m_1, m_2) = [a_0^4 - 4 t \beta(m_1, m_2)]^{1/4},
    \label{eq:a_over_time_circ}
\end{equation}
where $a_0$ is the initial semi-major axis. Moreover, we can solve for the merger time, the time until the binary will merge, by setting the final semi-major axis in Eq.~\ref{eq:a_over_time_circ}
\begin{equation}{\lwFuncURL{evol}{get_t_merge_circ}}
    t_{\rm merge, circ} = \frac{a_0^4}{4 \beta}.
    \label{eq:t_merge_circular}
\end{equation}

Inverting this function and applying the fact that we know that $e \to 0$ when the binary merges gives the merger time \citep[][Eq.~5.14]{Peters+1964}
\begin{equation}{\lwFuncURL{evol}{get_t_merge_ecc}}
    t_{\rm merge} = \frac{12}{19} \frac{c_{0}^{4}}{\beta} \int_0^{e_0} \frac{\left[1+(121 / 304) e^{2}\right]^{\frac{1181}{2299}}}{e^{-29 / 19}\left(1-e^{2}\right)^{3 / 2}} \mathrm{d}e.
    \label{eq:t_merge_eccentric}
\end{equation}
For very small or very large eccentricities we approximate this integral using the following expressions (given in unlabelled equations after \citealp[][Eq.~5.14]{Peters+1964})
\begin{align}
    t_{\rm merge,\, e^2 \ll 1} &= \frac{c_0^4}{4 \beta} \cdot e_0^{48 / 19}, \\
    t_{\rm merge,\, (1 - e^2) \ll 1} &= \frac{768}{425} \frac{a_0^4}{4 \beta} (1 - e_0^2)^{7/2}.
\end{align}
\noindent The standard threshold employed by \lw{} for small eccentricities is $e=0.01$ and for large eccentricities is $e=0.99$, though we note that this can be customised by the user as well.

Therefore, to account for the integration of the signal over the mission, we can introduce the `characteristic' strain amplitude of the $n^{\rm th}$ harmonic \citep[e.g][]{Finn+2000, Moore+2015}\footnote{Note that this is factor of 2 different from \citet{Finn+2000}. This is because the factor of 2 is already included in the \citet{Robson+2019} sensitivity curve and so is removed here.}
\begin{equation}{}
    h_{c, n}^2 = \qty(\frac{f_n^2}{\dot{f}_n}) h_{0, n}^2,
    \label{eq:strain-charstrain}
\end{equation}
where $h_{0, n}$ is the strain amplitude in the $n^{\rm th}$ harmonic.

\subsubsection{Strain}
In order to obtain an expression for the strain amplitude of gravitational waves in the $n^{\rm th}$ harmonic, we can use Eq.~\ref{eq:strain-charstrain} and plug in Eq.~\ref{eq:fdot} and Eq.~\ref{eq:char_strain} 
\begin{equation}{\lwFuncURL{strain}{h_0_n}}
    h_n^2 = \frac{2^{28/3}}{5} \frac{(G \mathcal{M}_c)^{10/3}}{c^8 D_L^2} \frac{g(n, e)}{n^2} \qty(\pi f_{\rm orb})^{4/3} . 
    \label{eq:strain}
\end{equation}

\subsubsection{Amplitude modulation for orbit averaged sources}
\label{sec:amplitude_mod}
Because the LISA detectors are not stationary and instead follow an Earth-trailing orbit, the antenna pattern of LISA is not isotropically distributed or stationary. For sources that have a known position, inclination, and polarisation, we can consider the amplitude modulation of the strain due to the average motion of LISA's orbit. We closely follow the results of \citet{Cornish2003} to write down the amplitude modulation as

\begin{equation}{\lwFuncURL{strain}{amplitude_modulation}}
    A_{\rm{mod}}^{2}=\frac{1}{2} \left[\left(1+\cos ^{2} \iota\right)^{2}\left\langle F_{+}^{2}\right\rangle+4 \cos ^{2} \iota\left\langle F_{\times}^{2}\right\rangle\right],
\label{eq:amp_mod}
\end{equation}
where $\left\langle F_{+}^{2}\right\rangle$ and $\left\langle F_{\times}^{2}\right\rangle$, the orbit-averaged detector responses, are defined as
\begin{equation}{\lwFuncURL{utils}{F_plus_squared}}
    \begin{split}
        \avg{ F_{+}^{2}} = \frac{1}{4}\big(\cos ^{2} 2 \psi\left\langle D_{+}^{2}\right\rangle &-\sin 4 \psi\left\langle D_{+} D_{\times}\right\rangle\\
        &+\sin ^{2} 2 \psi\left\langle D_{\times}^{2}\right\rangle\big),
    \end{split}
\label{eq:response_fplus}
\end{equation}

\begin{equation}{\lwFuncURL{utils}{F_cross_squared}}
    \begin{split}
        \avg{ F_{\times}^{2}} = \frac{1}{4}\big(\cos^{2} 2 \psi \avg{ D_{\times}^{2}} &+\sin 4 \psi \avg{ D_{+} D_{\times}}\\
        &+\sin ^{2} 2 \psi \avg{ D_{+}^{2}} \big),
    \end{split}
\label{eq:response_fcross}
\end{equation}
and
\begin{equation}{\lwFuncURL{utils}{D_plus_D_cross}}
    \begin{split}
       \avg{ D_{+} D_{\times}} = \frac{243}{512} \cos \theta \sin 2 \phi &\left(2 \cos ^{2} \phi-1\right) \\ &\times \left(1+\cos ^{2} \theta\right),
    \end{split}
\label{eq:d_plus_cross}
\end{equation}
\begin{equation}{\lwFuncURL{utils}{D_cross_squared}}
    \begin{split}
        \avg{ D_{\times}^{2}} = \frac{3}{512}\big(120 \sin ^{2} \theta &+\cos ^{2} \theta \\
        &+162 \sin ^{2} 2 \phi \cos ^{2} \theta\big),
    \end{split}
\label{eq:d_cross}
\end{equation}
\begin{equation}{\lwFuncURL{utils}{D_plus_squared}}
    \begin{split}
       \avg{ D_{+}^{2} } =  \frac{3}{2048}\big[&487+158 \cos ^{2} \theta+7 \cos ^{4} \theta\\
        &-162 \sin ^{2} 2 \phi\left(1+\cos ^{2} \theta\right)^{2}\big].
    \end{split}
\label{eq:d_plus}
\end{equation}
In the equations above, the inclination is given by $\iota$, the right ascension and declination are given by $\phi$ and $\theta$ respsectively, and the polarisation is given by $\psi$.

The orbital motion of LISA smears the source frequency by roughly $10^{-4}\,\rm{mHz}$ due to the antenna pattern changing as the detector orbits, the Doppler shift from the motion, and the phase modulation from the $+$ and $\cross$ polarisations in the antenna pattern. Generally, the modulation reduces the strain amplitude because the smearing in frequency reduces the amount of signal build up at the true source frequency.

We note that since the orbit averaging is carried out in Fourier space, this requires the frequency to be monochromatic and thus is only implemented in \lw{} for quasi-circular binaries. We also note that since the majority of the calculations in \lw{} are carried out for the full position, polarisation, and inclination averages, we place a pre-factor of $5/4$ on the amplitude modulation in the software implementation to undo the factor of $4/5$ which arises from the averaging of Equations~\ref{eq:response_fplus} and \ref{eq:response_fcross}.

We note that this section draws heavily from \citet{Flanagan+1998} Section II C.

In LISA's case, when averaged over all angles and polarisations, the antenna patterns are orthogonal and thus $\langle F_+ F_{\times}\rangle = 0$. This means we can rewrite Eq.~\ref{eq:signal} as 
\begin{equation}{}
    \begin{split}
        |\tilde{s}(f)|^2 &= |F_+(\theta, \phi, \psi)\tilde{h}_+(t, D_L, \iota, \beta)|^2\\
        &+ |F_{\times}(\theta, \phi, \psi)\tilde{h}_{\times}(t, D_L, \iota, \beta)|^2,
    \end{split}
\end{equation} 
which can then be applied to Eq.~\ref{eq:snr_general_start} as
\begin{equation}{}
\label{eq:snr_general_simpler}
    \rho^2 = 4 \int_0^{\infty} \frac{|F_+\tilde{h}_+|^2 + |F_{\times}\tilde{h}_{\times}|^2}{P_{\rm n}(f)} df.
\end{equation}

\begin{align}
    \int_{-\infty}^{+\infty}dt\int dA \frac{dE}{dAdt} &= \int_{0}^{\infty}df \int dA \frac{\pi f^2}{2} \left(|\tilde{h}_{+}|^2 + |\tilde{h}_{\times}|^2 \right).
\label{eq:Parseval}
\end{align}
Note that the factor of frequency squared comes from the Fourier transform of the square of the time derivative in Eq.~\ref{eq:Parseval}. Now, since $A = D_L^2 \Omega$ and $|\tilde{h}_{+,\times}|^2 = |\tilde{H}_{+,\times}|^2 / D_L^2$, we find
\begin{equation}{}
\label{eq:little_h_to_big}
    |\tilde{h}_{+,\times}|^2 dA = |\tilde{H}_{+,\times}|^2 d\Omega_{\iota,\beta},
\end{equation}
then we can write Eq.~\ref{eq:Parseval} in terms of $|H_{+,\times}|^2$ as 
\begin{equation}{}
    \int_{-\infty}^{+\infty}dt\int dA \frac{dE}{dAdt} = \int_{0}^{\infty}df \frac{\pi f^2}{2} \int d\Omega \left(|\tilde{H}_{+}|^2 + |\tilde{H}_{\times}|^2 \right).
\label{eq:Parseval_2}
\end{equation}
Alternatively, by using Eq.~\ref{eq:little_h_to_big} and performing a Fourier transform we can also write that
\begin{equation}{}
    \int_{-\infty}^{+\infty}dt\int dA \frac{dE}{dAdt} = \int_{0}^{\infty}df \int d\Omega \frac{dE}{d\Omega df}.
    \label{eq:deriv_relation}
\end{equation}
From inspection of Eq.~\ref{eq:Parseval_2} and Eq.~\ref{eq:deriv_relation}, we can write the spectral energy flux as
\begin{equation}{}
    \int d\Omega \frac{dE}{d\Omega df} = \frac{\pi f^2}{2} \int d\Omega \left(|\tilde{H}_{+}|^2 + |\tilde{H}_{\times}|^2 \right) .
    \label{eq:se_flux}
\end{equation}

For SNR calculations that take into account the amplitude modulation due to LISA's orbital motion (Section~\ref{sec:amplitude_mod}, we apply the calculations as described above but include the modulation in either the strain or characteristic strain as is appropriate.

@@

Let us compute the horizon distance for a grid of chirp masses and orbital frequencies. First, we can recall that the signal-to-noise ratio of a source is inversely proportional to its distance from LISA and so we can find the horizon distance, $D_{\rm hor}$ as
\begin{equation}{}
    \label{eq:snr_to_hor_dist}
    D_{\rm hor} = \frac{\rho(D)}{\rho_{\rm detect}} \cdot D,
\end{equation}
where $\rho(D)$ is the signal-to-noise ratio as some distance $D$ and $\rho_{\rm detect}$ is the threshold above which we consider a source to be detectable. For the purpose of this example we will set $\rho_{\rm detect} = 7$. We can then, as a function of chirp mass and frequency, determine the maximum distance for which the source of interest is detected based on this SNR threshold.

This can be most efficiently accomplished using \lw{}'s source class. Although the source class only reads in one-dimensional lists of sources, one can flatten a grid of chirp masses and orbital frequencies to make the input work. \lw{} can then compute the merger times and signal-to-noise ratio of each source with only two lines of code. These results can then be reshaped to match the original grid of chirp masses and frequencies.

We convert these SNRs to horizon distances using Eq.~\ref{eq:snr_to_hor_dist} and plot the result in Figure~\ref{fig:horizon_distance}. The shape of the LISA sensitivity curve is clearly reflected in the horizon distance. This is because circular sources only emit at a single frequency and thus every feature in the sensitivity curve has a strong effect on the SNR (and therefore horizon distance) for circular sources. We also see that once the merger time of a source is shorter than the LISA mission length (shown by the black dotted line), the horizon distance sharply decreases. This is because if a source merges before the LISA mission concludes, it has less time to accumulate signal and thus has a lower SNR and horizon distance.

@@

We see two effects in the signal-to-noise ratio here. First, increasing the eccentricity from essentially circular to $e = 0.6$ results in a higher signal-to-noise ratio ($\rho=31.7 \to \rho=50.2$). This is because an eccentric binary has enhanced energy emission via gravitational waves \citep{Peters+1963}. This means that an eccentric binary will not only inspiral faster than an otherwise identical circular binary, but also will always produce a stronger gravitational wave strain. The enhancement factor and its exact dependence on eccentricity is discussed in more detail Section~\ref{sec:derivations} (specifically Eq.~\ref{eq:eccentricity_enhancement_factor}).

@@

\begin{figure}[htb]
    \centering
    \includegraphics[width=\columnwidth]{figures/detector_sc_compare.pdf}
    \caption{The strain spectral density of the LISA detector with different specifications \citep{Robson+2019} and the TianQin detector \citep{Huang+2020}. We show the LISA curve for three different mission lengths and once with an approximate response function.}
    \label{fig:detector_sc_compare}
\end{figure}

@@

Note that \lw{} could also be used in this way to find the SNR of any system in which the orbital evolution is known. Thus for a triple system or a binary experiencing gas drag (as long the evolution of the eccentricity and orbital frequency is known) \lw{} is entirely capable of calculating the SNR evolution.

@@

\subsection{LISA verification binaries\texorpdfstring{ - \tutorialLink{https://legwork.readthedocs.io/en/latest/demos/VerificationBinaries.html}}{}}
\citet{Kupfer+2018} presents the LISA verification binaries, a collection of known binary systems that have gravitational-waves that are strong enough to be detected by LISA. In \lw{}, we provide easy access to this data through the \href{https://legwork.readthedocs.io/en/latest/api/legwork.source.VerificationBinaries}{\color{\lwColour}{\texttt{VerificationBinaries}}} class, which is a subclass of \href{https://legwork.readthedocs.io/en/latest/api/legwork.source.Source}{\color{\lwColour}{\texttt{Source}}}. This means that the class works identically to \texttt{source} but has the verification binary data preloaded into the variables. In addition to the base variables, this class also includes the designation of each binary and the SNR that \citet{Kupfer+2018} computed\footnote{We note that this SNR differs from the SNR that \lw{} gives for each binary. This is because \citet{Kupfer+2018} run a full LISA simulation for these binaries, whilst we follow the orbit-averaged approach (see Section~\ref{sec:amplitude_mod}), since the former would be intractable for any large number of sources.}.

As an example of how you could use this data, we use the \texttt{VerificationBinaries} class to plot the sources on the LISA sensitivity curve with the SNR calculated by \citet{Kupfer+2018} for a 4-year LISA mission and colour the binaries by their primary mass. We see that the verification binaries tend to be detected between $1$-$10 \unit{mHz}$ and have masses estimated to be between $0.1$-$1 \unit{M_{\odot}}$.

\begin{figure}[htb]
    \centering
    \includegraphics[width=\columnwidth]{verification_binaries_on_sc.pdf}
    \caption{The LISA verification binaries from \citet{Kupfer+2018} plotted on the LISA sensitivity curve \citep{Robson+2019} for a 4-year LISA mission. Points are coloured by their primary mass.}
    \label{fig:verification_binaries_on_sc}
\end{figure}

\section{Conclusion \& Summary}\label{sec:summary}
We have presented \lw{}, a package designed to aid in calculations for stellar-origin binary sources of mHz gravitational-wave observatories like LISA. We outlined the implementation of orbital evolution due to gravitational-wave emission, gravitational wave strain, SNR, and visualisation modules and provided a detailed derivation for each of the equations required for each module. Finally, we provided several use case examples for how \lw{} can be used to better understand the detectability of compact-object binaries.
